### 2023-03-02 calculus.tex
Remove analytic projection rules
Cut down products section

% TODO standardize naming: marking vs mark insertion, marks vs error marks vs markings
% TODO uniform style for referencing rules

@@

\subsubsection{Holes}
For completeness, we finish the development of the marked lambda calculus on the GTLC with the
marking of holes. Like number literals and booleans, these are governed by straightforward rules.
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\EEHole}{\TUnknown}
  }{USHole}

@@

We now demonstrate how one might extend or adapt the marked lambda calculus for new typing features
by adding binary product types. First, in \cref{fig:calculus-product-judgments}, we extend the
syntax of types and the accompanying notions of consistency and join, which closely mirror the
definitions for arrow types.
%
\[\begin{array}{rrcl}
  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TProd{\TMV}{\TMV}
\end{array}\]

We additionally define an auxiliary notion of \emph{matched binary product types}: the judgment
$\matchedProd{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be considered the binary product
type $\TProd{\TMV_1}{\TMV_2}$.

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\consistent{\TMV_1}{\TMV_2}}} $\TMV_1$ is consistent with $\TMV_2$
  %
  \begin{mathpar}
    \judgment{
      \consistent{\TMV_1}{\TMV_1'} \\
      \consistent{\TMV_2}{\TMV_2'} \\
    }{
      \consistent{\TProd{\TMV_1}{\TMV_2}}{\TProd{\TMV_1'}{\TMV_2'}}
    }{TCProd} \\
  \end{mathpar}

  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
  %
  \[\begin{array}{rcl}
    & \cdots & \\
    \joinsTo{(\TProd{\TMV_1}{\TMV_2})}{(\TProd{\TMV_1'}{\TMV_2'})}{\TProd{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
  \end{array}\]

  \judgbox{\ensuremath{\matchedProd{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched binary product type $\TProd{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedProd{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMPUnknown}

    \judgment{ }{
      \matchedProd{\TProd{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
    }{TMPProd}
  \end{mathpar}
  %
  \caption{Binary product consistency, join, and matched product type.}
  \label{fig:calculus-product-judgments}
\end{figure}

The unmarked language is extended with a pair constructor and projection operators. In addition to
the synthesis rules, we have explicit analysis rules. If an expected type is known for a pair
expression, the responsibility for type errors is given to the components. Likewise, we shift blame
to the projected expression when analysing projections.
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots
                               \mid \EPair{\EMV}{\EMV}
                               \mid \EProjL{\EMV} \mid \EProjR{\EMV}
\end{array}\]
%
\begin{mathpar}
  \judgment{
    \ctxSynTypeU{\ctx}{\EMV_1}{\TMV_1} \\
    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TProd{\TMV_1}{\TMV_2}}
  }{USPair}

  \judgment{
    \ctxSynTypeU{\ctx}{\EMV}{\TMV} \\
    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\EProjL{\EMV}}{\TMV_1}
  }{USProjL}

  \judgment{
    \ctxSynTypeU{\ctx}{\EMV}{\TMV} \\
    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\EProjR{\EMV}}{\TMV_2}
  }{USProjR} \\

  \judgment{
    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TMV_1} \\
    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV_2}
  }{
    \ctxAnaTypeU{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TMV}
  }{UAPair}

  \judgment{
    \ctxAnaTypeU{\ctx}{\EMV}{\TProd{\TMV}{\TUnknown}}
  }{
    \ctxAnaTypeU{\ctx}{\EProjL{\EMV}}{\TMV}
  }{UAProjL}

  \judgment{
    \ctxAnaTypeU{\ctx}{\EMV}{\TProd{\TUnknown}{\TMV}}
  }{
    \ctxAnaTypeU{\ctx}{\EProjR{\EMV}}{\TMV}
  }{UAProjR}
\end{mathpar}
%
Similarly to application, \textsc{USProjL} and \textsc{USProjR} employ the matched binary product
type judgment; a projection $\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may
be considered to be a pair.

The marked language is extended with the same forms, and we devise the following marking rules,
again modelling off the above unmarked typing rules:
%
\[\begin{array}{rrcl}
  \ECMName & \ECMV & \Coloneqq & \cdots
                               \mid \ECPair{\ECMV}{\ECMV}
                               \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                               \mid \ECSynNonMatchedProd{\ECMV}
\end{array}\]
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
  }{MKSPair}

@@

  \judgment{
    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MKAPair1}

  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TProd{\TMV}{\TUnknown}}
  }{
    \ctxAnaFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV}
  }{MKAProjL}

  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TProd{\TMV}{\TUnknown}}
  }{
    \ctxAnaFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV}
  }{MKAProjR}
\end{mathpar}
%
accompanied by corresponding typing rules for marked expressions:
%
\begin{mathpar}
  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV_1}{\TMV_1} \\
    \ctxSynTypeM{\ctx}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynTypeM{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
  }{MSPair}

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\\\
    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynTypeM{\ctx}{\ECProjL{\ECMV}}{\TMV_1}
  }{MSProjL1}

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\\\
    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynTypeM{\ctx}{\ECProjR{\ECMV}}{\TMV_2}
  }{MSProjR1}

  \judgment{
    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TMV_1} \\\\
    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TMV_2}
  }{
    \ctxAnaTypeM{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MAPair1}

  \judgment{
    \ctxAnaTypeM{\ctx}{\ECMV}{\TProd{\TMV}{\TUnknown}}
  }{
    \ctxAnaTypeM{\ctx}{\ECProjL{\ECMV}}{\TMV}
  }{MAProjL}

  \judgment{
    \ctxAnaTypeM{\ctx}{\ECMV}{\TProd{\TUnknown}{\TMV}}
  }{
    \ctxAnaTypeM{\ctx}{\ECProjR{\ECMV}}{\TMV}
  }{MAProjR}
\end{mathpar}

However, as previously seen with lambda abstractions and applications, two sets of rules are further
necessary to satisfy totality. Firstly, the premise in \textsc{UAPair} that $\TMV$ is a matched
binary product may fail, requiring marking and typing rules to handle that case:
%
\begin{mathpar}
  \judgment{
    \notMatchedProd{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TUnknown} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MKAPair2}

  \judgment{
    \notMatchedProd{\TMV} \\
    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TUnknown} \\
    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TUnknown}
  }{
    \ctxAnaTypeM{\ctx}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MAPair2}
\end{mathpar}

Another kind of error may arise when the subject of a projection does not synthesize a matched
product type. As in our development of marking for applications, we utilize a new error mark
$\ECSynNonMatchedProd{\ECMV}$, leading to the following rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
    \notMatchedProd{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjLSynNonMatchedProd{\ECMV}}{\TMV_1}
  }{MKSProjL2}

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
    \notMatchedProd{\TMV}
  }{
    \ctxSynTypeM{\ctx}{\ECProjLSynNonMatchedProd{\ECMV}}{\TUnknown}
  }{MSProjL2} \\

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
    \notMatchedProd{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjRSynNonMatchedProd{\ECMV}}{\TMV_2}
  }{MKSProjR2}

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
    \notMatchedProd{\TMV}
  }{
    \ctxSynTypeM{\ctx}{\ECProjRSynNonMatchedProd{\ECMV}}{\TUnknown}
  }{MSProjR2}
\end{mathpar}

>>>>

@@

\subsubsection{Holes}
For completeness, we finish the development of the marked lambda calculus on the extended GTLC with
the marking of holes. Like number literals and booleans, these are governed by straightforward
rules.
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\EEHole}{\TUnknown}
  }{USHole}

@@

We now demonstrate how one might extend or adapt the marked lambda calculus for new typing features
by adding binary product types, written $\TProd{\TMV}{\TMV}$. In the interest of space, we focus on
aspects related to marking; a comprehensive set of rules is found in the supplemental material.

First, we extend the syntax of types and the accompanying notions of consistency and join, which
closely mirror the definitions for arrow types. Additionally, analogous to the notion of matched
arrow types, the judgment $\matchedProd{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be
considered the binary product type $\TProd{\TMV_1}{\TMV_2}$ [cite?].

The unmarked language is extended with a pair constructor $\EPair{\EMV}{\EMV}$ and projection
operators $\EProjL{\EMV}$ and $\EProjR{\EMV}$, governed by standard bidirectional rules. In addition
to the synthesis rules, an explicit analysis rules is defined for the former: if an expected type is
known, the responsibility for type errors is shift to the components.

The marked language is extended with the same forms, and we again devise marking rules by modelling
off the unmarked typing rules.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
  }{MKSPair}

@@

  \judgment{
    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MKAPair1}
\end{mathpar}

However, as previously seen with lambda abstractions and applications, two sets of rules are further
necessary to satisfy totality. Firstly, the premise in \textsc{UAPair} that $\TMV$ is a matched
binary product may fail, requiring the introduction of a new error mark,
$\ECAnaNonMatchedProd{\ECMV}$, which is analogous to $\ECAnaNonMatchedArrow{\ECMV}$ for lambda
abstractions. Another kind of error may arise when the subject of a projection does not synthesize a
matched product type. As in our development of marking for applications, we utilize a new mark,
written $\ECSynNonMatchedProd{\ECMV}$. These cases give the final marking rules for products:
%
\begin{mathpar}
  \judgment{
    \notMatchedProd{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TUnknown} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MKAPair2} \\

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
    \notMatchedProd{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjLSynNonMatchedProd{\ECMV}}{\TMV_1}
  }{MKSProjL2}

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
    \notMatchedProd{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjRSynNonMatchedProd{\ECMV}}{\TMV_2}
  }{MKSProjR2}
\end{mathpar}
